Reducing Acceptance Marks in Emerson-Lei Automata by QBF-solving

### 2022-04-26 paper/main.tex
progress in section 6

If some call of the function
$\mathit{satisfiable}(\Phi^\mathit{opt}_{j,C,K{-}1})$ in the first
\textbf{while} loop of Algorithm~\ref{alg:single} does not return
$\true$, then the name of the output automaton encodes the reason for
it. In case of a single level reduction, the name has the form
\texttt{L$j$\_$k$\_$X$}, where $j$ is the considered level, $k=K-1$ is
the number of acceptance mark considered by the formula, $X$ is either
\texttt{U} if the formula is unsatisfiable or \texttt{T} if the solver
ran out of resources. If $X$ is \texttt{T}, we know that additional
resources (typically longer timeout) may lead to further reductions.
If the multi-level reduction is used, the automaton name contains the
information from all levels. For example, the name `\texttt{L1\_5\_U
  L2\_3\_U L3\_1\_T}' means that level 1 reduced the number of marks
to 6 (reduction to 5 is impossible on this level), level 2 reduced it
to 4, and level 3 to 2 and the automaton may be potentially reduced
further.

@@

We have used the tool \texttt{genltl} of the Spot
library~\cite{duret.16.atva2} to generate LTL formulas from
literature. The parametrized formulas were generated for all
combinations of parameter values from 1 to 4. Altogether, we have
generated 348 unique LTL formulas. These have been translated to
deterministic TELA by two state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva} and to nondeterministic TELA by
another two state-of-the-art translators, namely \texttt{ltl2tgba}
from the Spot library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}. Each of the four automata sets
produced by individual tools contain less than 348 automata as not all
of the formulas are successfully translated due to timeout of 60
seconds. Further, we have removed automata with 0 or 1 acceptance
marks as there is little point reducing these. Table
\ref{tab:ltltranslators} shows the exact versions of used translators,
the homepage of the used tools and the number of automata produced by
the given tool from the LTL formulas after filtering.

We have used multiple other helper tools in the experimental
evaluation.  Alongside \texttt{genltl}, we have used \texttt{ltlfilt}
to simplify the formulas and to remove duplicates, and
\texttt{autcross} to compare automata and measure the running times.

Spot version 2.10.3.dev and Z3 version 4.8.15 have been used in the
experimental evaluation. The experiments have been run on a computer
with Intel® Core™ i7-8700 processor and 32 GB of memory running
Ubuntu 20.04.4.

For each automata set (that is, the output of an individual LTL
translator), we have applied all single-level reductions and the
multi-level reduction, always with incremental approach and without
reduction of the number of clauses. The timeout for each QBF query was
set to 30 seconds.
%We have again filtered out the automata for which
%any of the four \telatko runs timed out after 120 seconds. We do this
%to make sure the numbers of acceptance marks can be meaningfully
%compared between different reduction levels of the same automata set.
%Additionally, we have filtered one automaton from the \texttt{ltl3tela} set
%that cannot be processed by \telatko due to an extensive use of system
%resources.

Table~\ref{tab:cumm-data} shows the cummulative numbers of marks in
the input automata sets and after each reduction, together with the
reduction ratio and total time spent by the considered reduction. The
columns \textit{solver response} and \textit{solver timeout} show the
number of automata for which the last query to QBF solver has been
successfully answered, or the solver timed out within a 30 seconds
limit to a single query, respectively. The timeout of the last QBF
query implies that the automaton could be potentially further reduced
by the reduction with a longer time limit.

@@

\begin{table}[h]
\caption{The cummulative number of acceptance marks before and after reduction
for various automata sets.
%\textit{X/Y solved aut.} means that out of
%\textit{Y} successfully translated LTL formulas, every run of \telatko ended
%within 120 seconds time limit on \textit{X} automata. We sum up the acceptance
%marks only on these automata.
The columns \textit{solver response} and
\textit{solver timeout} refer to last call to the QBF solver (that is, the
numbers of its successful responses within 30 seconds limit and timeouts).
}
\label{tab:cumm-data}
\centering
\input{table-data.tex}
\end{table}

Table~\ref{tab:heatmap} shows the number of automata coming from all
four sets that have a given number of original acceptance marks and a
given number of acceptance marks after multi-level reduction. The
table indicates that in many cases only 1 or 2 marks can be
saved. However, the achieved reduction is substantial for some
automata with a higher number of original acceptance marks. For
example, in 22 cases, we have successfully reduced 7 or more
acceptance marks to only 4.
% Moreover, there were many timeouts of the last QBF query for manof these automata.
% Let us note that some that some reductions are the results are promising, e.g. in 17 cases, we have successfully
% reduced 8 or more acceptance marks to only 4.

@@

\begin{figure}[h]
\caption{Performance of \telatko on different automata sets. The $x$ axis
represents the number of reduced automata, the $y$ axis shows the time spent
reducing (in seconds).}
\label{fig:quantile}
\centering
\input{quantile.tex}
\end{figure}

@@

\section{TODO}

Do submission:
\begin{itemize}
\item upravit nazev a abstrakt, zduraznit uzitecnost TELA pro concurrency (analysis of probabilistic systems...)
\item do experimentu napsat, jaky hardware pouzivame (merit cas bez toho nedava smysl)
\item spustit experimenty bez timeoutu na telatko (?a delsi timeout na QBF?)
\item prelestit text v poslednich dvou kapitolach
\end{itemize}

\noindent
Do budoucna:
\begin{itemize}
\item level 1 bez SCC optimalizace
\item nahradit weak SCC jednou hranou pro neakceptujici a jinou pro akceptujici
\item sladit konstrukci QBF formuli s implementaci
\item podivat se znovu na uzitecnost a potrebnost optimalizaci typu ``klauzule neni false'' a ``alespon jedna klauzule je neprazdna''
\item sladit popis algoritmu s implementaci
\item v experimentech nejak sledovat vztah usetreneho casu a usetrenych znacek?
\item zkusit aplikovat na jine typy automatu (treba TGBA nebo Rabin)?
\item zkusit, zda/jak redukce znacek umozni redukovat stavovy prostor
\end{itemize}

>>>>

If some call of the function
$\mathit{satisfiable}(\Phi^\mathit{opt}_{j,C,K{-}1})$ in the first
\textbf{while} loop of Algorithm~\ref{alg:single} does not return
$\true$, then the name of the output automaton encodes the reason for
it. In case of a single level reduction, the name has the form
\texttt{L$j$\_$k$\_$X$}, where $j$ is the considered level, $k=K-1$ is
the number of acceptance mark considered by the formula, $X$ is either
\texttt{U} if the formula is unsatisfiable or \texttt{T} if the solver
did not decided within the time limit. If $X$ is \texttt{T}, we longer
timeout may lead to further reductions. a If the multi-level reduction
is used, the automaton name contains the information from all
levels. For example, the name `\texttt{L1\_5\_U L2\_3\_U L3\_1\_T}'
means that level 1 reduced the number of marks to 6 (reduction to 5 is
impossible on this level), level 2 reduced it to 4, and level 3 to 2
as the QBF solver did not finished in the time limit when trying to
reduce the number of marks to
1.%when tryint the automaton may be potentially reduced further.

@@

We have used the tool \texttt{genltl} of the Spot
library~\cite{duret.16.atva2} to generate LTL formulas from
literature. The parametrized formulas were generated for all
combinations of parameter values from 1 to 4. We used the tool
\texttt{ltlfilt} of the Spot library to simplify the formulas and
remove duplicates. Altogether, we have generated 348 unique LTL
formulas. These have been translated to deterministic TELA by two
state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva} and to nondeterministic TELA by
another two state-of-the-art translators, namely \texttt{ltl2tgba}
from the Spot library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}. Each of the four automata sets
produced by individual tools contain less than 348 automata as not all
of the formulas are successfully translated due to timeout of 60
seconds. Further, we have removed automata with 0 or 1 acceptance
marks as there is a little point reducing these. Table
\ref{tab:ltltranslators} shows the exact versions and homepages of
used translators, and the number of automata produced by each
translator after removing the automata with less than two marks.

@@

For each automata set (that is, the automata produced by an individual
LTL translator), we have applied all single-level reductions and the
multi-level reduction, always with incremental approach and without
reduction of the number of clauses. The timeout for each QBF query was
set to 30 seconds.

For the experiments we used the version of \texttt{telatko} build with
Spot library version 2.10.4 and Z3 version 4.8.15. The experiments
have been run on a computer with Intel® Core™ i7-8700 processor and 32
GB of memory running Ubuntu 20.04.4. We used the tool
\texttt{autcross} of the Spot library to get the statistics of the
reduced automata and the running times.

Table~\ref{tab:cumm-data} shows the cummulative numbers of marks in
the input automata sets and after each reduction, together with the
reduction ratio and total time spent by the considered reduction. The
column \textit{solver timeout} shows the number of automata for which
the last query to QBF solver did not finished within the 30 seconds
limit. The timeout of the last QBF query implies that the automaton
may be potentially further reduced by the reduction with a longer time
limit.

\begin{table}[t]
\caption{The cummulative numbers of acceptance marks before and after reduction
for various reductions and automata sets.
%\textit{X/Y solved aut.} means that out of
%\textit{Y} successfully translated LTL formulas, every run of \telatko ended
%within 120 seconds time limit on \textit{X} automata. We sum up the acceptance
%marks only on these automata.
The column \textit{solver timeout} indicates the number of instances
when the last call to the QBF solver timed out.  }
\label{tab:cumm-data}
\centering
\setlength{\tabcolsep}{10pt}
\input{table-data.tex}
\end{table}

Table~\ref{tab:heatmap} shows the number of automata coming from all
four sets that have a given number of original acceptance marks and a
given number of acceptance marks after multi-level reduction. The
table indicates that in many cases only 1 or 2 marks can be
saved. However, the achieved reduction is substantial for some
automata with a higher number of original acceptance marks. For
example, in 24 cases, we have successfully reduced 7 or more
acceptance marks to only 4.
% Moreover, there were many timeouts of the last QBF query for manof these automata.
% Let us note that some that some reductions are the results are promising, e.g. in 17 cases, we have successfully
% reduced 8 or more acceptance marks to only 4.

@@

\begin{figure}[b]
\caption{Performance of \telatko on different automata sets. The $x$ axis
represents the number of reduced automata, the $y$ axis shows the time spent
reducing (in seconds).}
\label{fig:quantile}
\centering
\input{quantile.tex}
\end{figure}

\todo{dodelat popis os u quantile plotu}

@@

% \section{TODO}

% \noindent
% Do budoucna:
% \begin{itemize}
% \item level 1 bez SCC optimalizace
% \item nahradit weak SCC jednou hranou pro neakceptujici a jinou pro akceptujici
% \item sladit konstrukci QBF formuli s implementaci
% \item podivat se znovu na uzitecnost a potrebnost optimalizaci typu ``klauzule neni false'' a ``alespon jedna klauzule je neprazdna''
% \item sladit popis algoritmu s implementaci
% \item v experimentech nejak sledovat vztah usetreneho casu a usetrenych znacek?
% \item zkusit aplikovat na jine typy automatu (treba TGBA nebo Rabin)?
% \item zkusit, zda/jak redukce znacek umozni redukovat stavovy prostor
% \end{itemize}
